Stability of virtual aggregations of prosumers through diversification

### 2015-07-30 Article.tex
spell check

The smart grid vision however goes further and revolutionizes this top-down centralized architecture by assuming that bidirectional power flows are allowed. This would change completely the nature of the traditional grid since production could be located down to the very end of the distribution networks. Nodes that were simply pure loads yesterday could behave tomorrow as generators or loads \cite{Ramchurn}. On the other hand, the penetration of renewables in the production is continuously growing, and is expected to become majority in a near future. These plants completely rely on the presence and the intensity of their respective ressources (wind, sun, tiles...). Balancing production and demand in such a scenario appears much more challenging and it seems inevitable that the grid should modernize its infrastructures to sustain this transition \cite{Europe}. The key to drive safely such a system is assumed to be information, and more precisely, the capacity to measure, communicate, and analyze data on real time.

@@

In this paper, we focus on these "\textit{nodes}" in the distribution network that can produce and consume electricity. More precisely, we consider a set of agents that own both renewable distributed energy ressources (DER) and electrical loads. The production of an agent can be, of course, used to meet its own demand, but in cases where it is over-producing, we consider that it has the possibility to inject, against financial compensation, its extra-production in the grid. Such an agent model is known as a "\textit{prosumer}" \cite{Rathnayaka2012} (and will be called accordingly in this paper). A key point of this work is to mix the interests of the grid operator with the ones of the prosumers in a single utility function. While the latter intend to maximize their benefits with higher production contracts, the operator primary concerns are related to the quality and stability of the injections. The coalitions that we wish to form should thus be both stable and productive as to satisfy both parts. This compromise can be difficult to obtain in real situations, especially when renewable generation and complex spatiotemporal correlation come into play.

@@

Forming coalitions/aggregations is an envisaged solution to both the number of entities soliciting the operator, and their high variabilities. Indeed, it has been argued that using a multi-level aggregators architecture for the control could lead to better performances on the communication side \cite{Negeri2012}, since an aggregator can be considered as a single node by the level above it. On the other hand, it is well-known that diversification of the assets is a way of minimizing risks when constructing a portfolio. One thus expect a more stable and predictable energy production for a coalition than for a single agent. Nevertheless, all coalitions are obviously not equally stable or productive, which means that special attention should be paid to the aggregation step. Recall that prosumers have both a consumption and a production component, each depending on location and time, meaning that there are complex underlying correlations between the agents. This is a central topic of the present paper : given N prosumers, what coalitions should be formed so that the compromise between expected production and variability is optimized ?

@@

As explained in the section above, allowing entities using mainly renewables to inject power in the grid is a difficult challenge. Indeed, on the contrary to fossil plants whose production can be scheduled in advance to meet the expected consumption, renewables by definition only produce when the ressource is present. Unfortunately, these moments do not necessarily coincide with the consumption peak hours \cite{Milligan2010}. One possible solution consists in shifting some of the loads towards high productivity periods. Coupled with demand side management techniques implemented on the end users' smart meters, this could result in less production spilled as well as smaller consumption peaks \cite{Milligan2010} \cite{Logenthiran2012}. Since end users tend to seek a maximum utility at a minimum cost, dynamic pricing is believed to be a good way to give incentives to them. By carefully scheduling the prices, it is assumed that the load curve can be shaped to some extend.

Dynamic pricing is an interesting and useful tool, but it has also some limitations. It is likely that dynamic pricing will serve as a shaping mass tool while finer techniques will be needed locally. A popular approach in this direction consists in deploying storage devices in the network and using them as electricity buffers. Basically these storages would be charged when there is a surplus of production, and discharged when the consumption exceeds the production. Although quite simple, this idea causes numerous challenges when it comes to its real implementation. As long as the system considered remains small, a centralized control of these buffers could be envisioned. However, for real large systems it is likely that more sophisticated decentralized algorithms will be necessary. In \cite{Zhang2013}, the authors introduce a distributed energy management system with a high penetration of renewables such that power is scheduled in a distributed fashion. In \cite{Shadmand2014} the optimal storage capacity problem is addressed. There is indeed an interesting tradeoff between the costs of the equipments and the expected availability of power. The authors develop a framework that enables them to exhibit a Pareto front of efficient solutions.

These are only a few possibilities for balancing production and demand in the smart grid. Most of the time, these techniques will be coupled with predictions of the upcoming load curves and weather conditions. Combining all these technologies enables aggregators to quantify their expected production and the inherent risk that comes with it. The optimization of expected returns to risk is a traditional goal in finance, and a wide literature exists on this topic. It is well-known for instance, that the more risk one is willing to take, the higher his potential gains. On the contrary, when investing exclusively on low risks assets, one should expect relatively small gains. This tradeoff is formalized in the Markowitz' portfolio theory \cite{Markowitz1959}. More precisely, given a set of assets for which we have some historic data of returns, the objective is to find a linear combination of these assets (the so-called portfolio) which maximizes the expected value while minimizing the variance of the portfolio's return. Markowitz's answer is a set of efficient portfolios that all optimize in some sense this tradeoff. If one is able to put a number on his risk acceptance or on the target expected return, the corresponding efficient portfolio is a priori the best option. One of the most controversial assumptions in the portfolio theory is that returns are jointly normally distributed (or, at least, that the return distribution is jointly elliptical). Some economist have pointed out the fact that this assumption might not capture well the reality of financial markets \cite{Chicheportiche2010}.

Nevertheless, one of the key point in the Markowitz theory is to consider explicitely the correlation between the assets since they impact directly the variances of the portfolios. Since the work of \cite{Mantegna1999}, an interresting approach consists in computing a distance metric based on the correlation coefficients in order to organize the series in a correlation graph. Nodes represent the series considered while the edges are weighted by the metric. Because the metric can be computed for all pairs, these graphs are complete and of little use as is. Historically, the approach used by \cite{Mantegna1999} was to compute a minimum spanning tree as to obtain a hierarchical clustering of the series. Later on, it was pointed out that, by definition, a spanning tree could not capture the underlying clustering structure hidden in the correlation graph. In this paper, we use another classical filtering technique called $ \epsilon $-graph \cite{Garas2008}. It consists in selecting a threshold $ \epsilon $, and filtering out edges with smaller weights. As we will see further in this paper, this approach has the advantage of preserving clusters of correlated series.

@@

In this paper, we use weather quantities like wind speed or solar irradiance as alternative data for generating realistic production and consumption series. Fortunately, these kinds of data are easier to find, and since the development of small personal weather stations, their geographical granularity keeps increasing.  Since these quantities depend both on time and location, we discretize time into slots and space into zones in the following. A zone is simply a portion of the considered region of study for which we sampled data. Therefore, if prosumers i and j are positioned on the same zone, they are exposed to the same weather. Adding some intra-zone noise can easily be done though not considered in this paper.

@@

For a prosumer i, it is possible to write both quantities as a sum over the distributed energy ressources ($ DER_{i} $) and loads ($ load_{i} $) of i :

@@

We denote by $ \nu_{i}(t) $ and $ \Psi_{i}(t) $ the wind speed (in $ m.s^{-1} $) and the solar irradiance ( in $ W.m^{-2} $) at the location of agent i and at time t, so that :
\begin{equation}
 P_{i}^{P}(t) = \sum_{k \in WT_{i}} \mathcal{F}_{WT}( \nu_{i}(t) ) + \sum_{k \in PV_{i} } \mathcal{F}_{PV}(\Psi_{i}(t) ) 
\end{equation}
Where $ \mathcal{F}_{WT} $ (resp. $ \mathcal{F}_{PV} $) is the power curve for the wind-turbines (resp. photovoltaic panels). We made here the implicit assumption that all wind-turbines (resp. photovoltaic panels) have the same power curve. The model can be easily extended to multiple power curves accounting for different types of generators. More details about power curves and their approximations can be found in the appendix and in \cite{Lydia2014}. The process for generating the $ P_{i} $ series is pictured in the first block of the process diagram (see figure \ref{fig:process}).

@@

Suppose now that coalition S has to suggest a production value $ P_{S}^{CRCT} $ to enter the market. This means that, during the time S is on the market, it will have to inject in the grid exactly $ P_{S}^{CRCT} $ at any time t and will be rewarded proportionally to this amount, with penalties if it deviates. Obviously, the actual extra-production will not be constant at this value and will oscillate due to intermittencies in the production and consumption. If S always produces more than $ P_{S}^{CRCT} $, it will never have to pay penalties, but it is losing some gains since it could have announced a higher contract value. If the production oscillates around $ P_{S}^{CRCT} $, by using batteries or demand side management techniques (see section \ref{sec:related}), S could be able to maintain its production to the contract value at any time. Nevertheless, if the oscillations are too important compared to the available storage capacity, S will probably break the contract and pay penalties. We can see that there is a return over risk tradeoff here, meaning that coalitions should find the right balance between announcing too low and losing some potential gains, and claiming too high and paying penalties.

@@

$ \rho_{ij} $ being the Pearson's correlation coefficient between $ P_{i} $ and $ P_{j} $. If the coalition $ \{i,j\}$ proposes a contract value $ P_{S}^{CRCT} $, all instants where $ \{i,j\}$ will produce less than $ P_{S}^{CRCT} $ is critical. Indeed, in this kind of situations, $ \{i,j\}$ will either have to discharge batteries to keep up with its contract, or pay penalities to the grid. The probability that $ \{i,j\}$ is underproducing compared to the contract : $ Pr[P_{i,j} \leq P^{CRCT}] $ is thus an important indicator of the coalition's quality. A well-known result for normal distributions is that the cumulative distribution function can be written as :
\begin{equation}
Pr[P_{ij} \leq P_{S}^{CRCT}] = \dfrac{1}{2} \left[ 1+ erf \left( \dfrac{P_{S}^{CRCT} - \mu_{ij}}{\sigma_{ij}\sqrt{2}} \right) \right] 
\end{equation}

@@

The amount of risks a given coalition is willing to take depends on a lot of things, among which its capacity to compensate for under-producing (using batteries, backup generators...), or the risk aversion of the aggregator. Selecting the right contract value appears thus as an interesting problem on its own that we plan to investigate in future works. In order to keep the present paper in a reasonable length, we simplify the contract value selection problem by giving some responsibilities to a third party named the grid operator. The role of the grid operator is to constrain the market entry to coalitions able to propose both sufficiently high and sufficiently credible contract values. More formally, let $ \phi \in [0,1] $ be the reliability threshold fixed by the grid operator as a maximum value for the probability of under-producing. The highest contract value that a coalition can propose is thus $ P_{S}^{CRCT \star} $ such that $ Pr[P_{ij} \leq P_{S}^{CRCT \star}] = \phi $. In the gaussian example, this translates by coalition $ \{i,j\}$ announcing :

@@

\begin{figure}
\includegraphics[scale=.6]{gaussian_P_star}
\caption{$ P_{S}^{CRCT \star} $ depending on reliability parameter $ \phi $ for gaussian distributions (see equation \ref{eq:P_star}). Blue curve with triangles stands for a coalition S with an expected production of 5 units and a standard deviation of 0.5. Under a grid policy of $ \phi = 0.1 $, it is able to announce a contract value of $ P_{S}^{CRCT \star} = 4.36 $. The same coalition in term of expected production ($\mu = 5 $), but with a higher variance ($ \sigma = 1.5 $, cyan curve with circles) can only afford a smaller contract value of $ P_{S}^{CRCT \star} = 3.07 $. The red curve with diamonds stands for a coalition with a higher expected production $(\mu = 7)$, but with a very high unpredictability ($ \sigma = 5 $). For low values of $ \phi $, this coalition is thus heavily penalized and can only afford a contract of $ 0.59 $ units. Under grid policy $ (\phi = 0.1, P^{MIN} = 2) $, this last coalition is thus not allowed to enter the market (red dot below the horizontal dashed line). }
\label{fig:Gaussian}
\end{figure}

@@

The gaussian assumption of this small example is convenient as it allows us to write $ P_{S}^{CRCT \star} $ analytically. Nevertheless, such assumption is rather unrealistic in practice. In the following, we keep the same framework but release this gaussian assumption unless the contrary is specified. This assumption will indeed be convenient for computing some parameter estimates.

@@

In order to get an estimator for $ \alpha $, we solve equation \ref{eq:derivative} in a gaussian case (as in section \ref{sec:notations}). Furthermore, since considering all the possible interactions between agents is analytically intractable, we use here a mean approximation. Any quantity x that varies over the agent set is thus simplified in its mean value $ \bar{x} $. Solving equation \ref{eq:derivative} for $ \alpha $ in these conditions yields :

@@

Before running the algorithms, we need to calibrate the utility function by choosing the value of the $ \alpha $ parameter. Recall that the purpose of this parameter is to take into account some constraints on the coalition's sizes if needed. In this paper, neither the communication network nor the electrical grid are explicitly considered. Thus, we do not have any technical constraints on coalition sizes even if we designed the utility such that these could be taken into account. We select the desired size as being $ \lfloor N/N_{COAL} \rfloor $ (where $ \lfloor.\rfloor $ means floor). Figure \ref{fig:real_utility2} shows how the mean utility of a coalition evolves with its size when the optimum size is set to 40 agents. Using equation \ref{eq:alpha_star} to estimate $ \alpha $ based on the mean quantities and gaussian approximations seems to give acceptable results for the utility function behavior.

@@

And we extend this measure to the coalition stuctures :

@@

We consider that prosumers fail randomly, and we denote by $ \psi \in [0,1] $ the fraction of agents that failed. Figure \ref{fig:resilience} exhibits how the resilience of the coalition structures evolve according to $ \psi $. On the top subplot, $ P^{MIN} $ was volontarily selected relatively low such that the resiliences of the three structures fit on the same figure. When the $ P^{MIN} $ requirement increases, the differences between the algorithms also increase as visible on the bottom subplot of figure \ref{fig:resilience}. The de-correlated coalitions seem to achieve a more resilient production on the market in the sense that they are able to sustain a higher fraction of node failures.

@@

Because in real situations, agents are prone to failure, resilience is also an important criterium for the quality of the aggregations. We therefore studied how the coalitions are able to remain on the market when their agents fail randomly one by one. We showed that, in this situation, the coalitions resulting from our algorithm better withstand losses of agents.

@@

\begin{itemize}
\item Average wind speed (in $m.s^{-1}$)
\item Nebulosity (integer in $[0,8] $)
\item Temperature (in degree celsius)
\end{itemize}

@@

Where $ \mathcal{F}_{i}^{heat}(\tau(t), t) $ is the power curve that maps the temperature to a heating consumption, and $ \mathcal{F}_{i}^{elec}(t) $ computes the consumption of agent i (other than heating) at a given hour of the day. In the simulation, all agents have a desired inside temperature $ T_{i} $, supposed to be a constant for simplification. By using termodynamic laws $ \mathcal{F}_{i}^{heat}(\tau(t), t) $ can be approximated by :

>>>>

The smart grid vision however goes further and revolutionizes this top-down centralized architecture by assuming that bidirectional power flows are allowed. This would change completely the nature of the traditional grid since production could be located down to the very end of the distribution networks. Nodes that were simply pure loads yesterday could behave tomorrow as generators or loads \cite{Ramchurn}. On the other hand, the penetration of renewables in the production is continuously growing, and is expected to become majority in a near future. These plants completely rely on the presence and the intensity of their respective resources (wind, sun, tiles...). Balancing production and demand in such a scenario appears much more challenging and it seems inevitable that the grid should modernize its infrastructures to sustain this transition \cite{Europe}. The key to drive safely such a system is assumed to be information, and more precisely, the capacity to measure, communicate, and analyze data on real time.

@@

In this paper, we focus on these "\textit{nodes}" in the distribution network that can produce and consume electricity. More precisely, we consider a set of agents that own both renewable distributed energy resources (DER) and electrical loads. The production of an agent can be, of course, used to meet its own demand, but in cases where it is over-producing, we consider that it has the possibility to inject, against financial compensation, its extra-production in the grid. Such an agent model is known as a "\textit{prosumer}" \cite{Rathnayaka2012} (and will be called accordingly in this paper). A key point of this work is to mix the interests of the grid operator with the ones of the prosumers in a single utility function. While the latter intend to maximize their benefits with higher production contracts, the operator primary concerns are related to the quality and stability of the injections. The coalitions that we wish to form should thus be both stable and productive as to satisfy both parts. This compromise can be difficult to obtain in real situations, especially when renewable generation and complex spatiotemporal correlation come into play.

@@

Forming coalitions/aggregations is an envisaged solution to both the number of entities soliciting the operator, and their high variability. Indeed, it has been argued that using a multi-level aggregators architecture for the control could lead to better performances on the communication side \cite{Negeri2012}, since an aggregator can be considered as a single node by the level above it. On the other hand, it is well-known that diversification of the assets is a way of minimizing risks when constructing a portfolio. One thus expect a more stable and predictable energy production for a coalition than for a single agent. Nevertheless, all coalitions are obviously not equally stable or productive, which means that special attention should be paid to the aggregation step. Recall that prosumers have both a consumption and a production component, each depending on location and time, meaning that there are complex underlying correlations between the agents. This is a central topic of the present paper : given N prosumers, what coalitions should be formed so that the compromise between expected production and variability is optimized ?

@@

As explained in the section above, allowing entities using mainly renewables to inject power in the grid is a difficult challenge. Indeed, on the contrary to fossil plants whose production can be scheduled in advance to meet the expected consumption, renewables by definition only produce when the resource is present. Unfortunately, these moments do not necessarily coincide with the consumption peak hours \cite{Milligan2010}. One possible solution consists in shifting some of the loads towards high productivity periods. Coupled with demand side management techniques implemented on the end users' smart meters, this could result in less production spilled as well as smaller consumption peaks \cite{Milligan2010} \cite{Logenthiran2012}. Since end users tend to seek a maximum utility at a minimum cost, dynamic pricing is believed to be a good way to give incentives to them. By carefully scheduling the prices, it is assumed that the load curve can be shaped to some extend.

Dynamic pricing is an interesting and useful tool, but it has also some limitations. It is likely that dynamic pricing will serve as a shaping mass tool while finer techniques will be needed locally. A popular approach in this direction consists in deploying storage devices in the network and using them as electricity buffers. Basically these storages would be charged when there is a surplus of production, and discharged when the consumption exceeds the production. Although quite simple, this idea causes numerous challenges when it comes to its real implementation. As long as the system considered remains small, a centralized control of these buffers could be envisioned. However, for real large systems it is likely that more sophisticated decentralized algorithms will be necessary. In \cite{Zhang2013}, the authors introduce a distributed energy management system with a high penetration of renewables such that power is scheduled in a distributed fashion. In \cite{Shadmand2014} the optimal storage capacity problem is addressed. There is indeed an interesting trade-off between the costs of the equipments and the expected availability of power. The authors develop a framework that enables them to exhibit a Pareto front of efficient solutions.

These are only a few possibilities for balancing production and demand in the smart grid. Most of the time, these techniques will be coupled with predictions of the upcoming load curves and weather conditions. Combining all these technologies enables aggregators to quantify their expected production and the inherent risk that comes with it. The optimization of expected returns to risk is a traditional goal in finance, and a wide literature exists on this topic. It is well-known for instance, that the more risk one is willing to take, the higher his potential gains. On the contrary, when investing exclusively on low risks assets, one should expect relatively small gains. This trade-off is formalized in the Markowitz' portfolio theory \cite{Markowitz1959}. More precisely, given a set of assets for which we have some historic data of returns, the objective is to find a linear combination of these assets (the so-called portfolio) which maximizes the expected value while minimizing the variance of the portfolio's return. Markowitz's answer is a set of efficient portfolios that all optimize in some sense this trade-off. If one is able to put a number on his risk acceptance or on the target expected return, the corresponding efficient portfolio is a priori the best option. One of the most controversial assumptions in the portfolio theory is that returns are jointly normally distributed (or, at least, that the return distribution is jointly elliptical). Some economist have pointed out the fact that this assumption might not capture well the reality of financial markets \cite{Chicheportiche2010}.

Nevertheless, one of the key point in the Markowitz theory is to consider explicitly the correlation between the assets since they impact directly the variances of the portfolios. Since the work of \cite{Mantegna1999}, an interesting approach consists in computing a distance metric based on the correlation coefficients in order to organize the series in a correlation graph. Nodes represent the series considered while the edges are weighted by the metric. Because the metric can be computed for all pairs, these graphs are complete and of little use as is. Historically, the approach used by \cite{Mantegna1999} was to compute a minimum spanning tree as to obtain a hierarchical clustering of the series. Later on, it was pointed out that, by definition, a spanning tree could not capture the underlying clustering structure hidden in the correlation graph. In this paper, we use another classical filtering technique called $ \epsilon $-graph \cite{Garas2008}. It consists in selecting a threshold $ \epsilon $, and filtering out edges with smaller weights. As we will see further in this paper, this approach has the advantage of preserving clusters of correlated series.

@@

In this paper, we use weather quantities like wind speed or solar radiance as alternative data for generating realistic production and consumption series. Fortunately, these kinds of data are easier to find, and since the development of small personal weather stations, their geographical granularity keeps increasing.  Since these quantities depend both on time and location, we discretize time into slots and space into zones in the following. A zone is simply a portion of the considered region of study for which we sampled data. Therefore, if prosumers i and j are positioned on the same zone, they are exposed to the same weather. Adding some intra-zone noise can easily be done though not considered in this paper.

@@

For a prosumer i, it is possible to write both quantities as a sum over the distributed energy resources ($ DER_{i} $) and loads ($ load_{i} $) of i :

@@

We denote by $ \nu_{i}(t) $ and $ \Psi_{i}(t) $ the wind speed (in $ m.s^{-1} $) and the solar radiance ( in $ W.m^{-2} $) at the location of agent i and at time t, so that :
\begin{equation}
 P_{i}^{P}(t) = \sum_{k \in WT_{i}} \mathcal{F}_{WT}( \nu_{i}(t) ) + \sum_{k \in PV_{i} } \mathcal{F}_{PV}(\Psi_{i}(t) ) 
\end{equation}
Where $ \mathcal{F}_{WT} $ (resp. $ \mathcal{F}_{PV} $) is the power curve for the wind-turbines (resp. photovoltaic panels). We made here the implicit assumption that all wind-turbines (resp. photovoltaic panels) have the same power curve. The model can be easily extended to multiple power curves accounting for different types of generators. More details about power curves and their approximations can be found in the appendix and in \cite{Lydia2014}. The process for generating the $ P_{i} $ series is pictured in the first block of the process diagram (see figure \ref{fig:process}).

@@

Suppose now that coalition S has to suggest a production value $ P_{S}^{CRCT} $ to enter the market. This means that, during the time S is on the market, it will have to inject in the grid exactly $ P_{S}^{CRCT} $ at any time t and will be rewarded proportionally to this amount, with penalties if it deviates. Obviously, the actual extra-production will not be constant at this value and will oscillate due to intermittences in the production and consumption. If S always produces more than $ P_{S}^{CRCT} $, it will never have to pay penalties, but it is losing some gains since it could have announced a higher contract value. If the production oscillates around $ P_{S}^{CRCT} $, by using batteries or demand side management techniques (see section \ref{sec:related}), S could be able to maintain its production to the contract value at any time. Nevertheless, if the oscillations are too important compared to the available storage capacity, S will probably break the contract and pay penalties. We can see that there is a return over risk trade-off here, meaning that coalitions should find the right balance between announcing too low and losing some potential gains, and claiming too high and paying penalties.

@@

$ \rho_{ij} $ being the Pearson's correlation coefficient between $ P_{i} $ and $ P_{j} $. If the coalition $ \{i,j\}$ proposes a contract value $ P_{S}^{CRCT} $, all instants where $ \{i,j\}$ will produce less than $ P_{S}^{CRCT} $ is critical. Indeed, in this kind of situations, $ \{i,j\}$ will either have to discharge batteries to keep up with its contract, or pay penalties to the grid. The probability that $ \{i,j\}$ is under-producing compared to the contract : $ Pr[P_{i,j} \leq P^{CRCT}] $ is thus an important indicator of the coalition's quality. A well-known result for normal distributions is that the cumulative distribution function can be written as :
\begin{equation}
Pr[P_{ij} \leq P_{S}^{CRCT}] = \dfrac{1}{2} \left[ 1+ erf \left( \dfrac{P_{S}^{CRCT} - \mu_{ij}}{\sigma_{ij}\sqrt{2}} \right) \right] 
\end{equation}

@@

The amount of risks a given coalition is willing to take depends on a lot of things, among which its capacity to compensate for under-producing (using batteries, backup generators...), or the risk aversion of the aggregator. Selecting the right contract value appears thus as an interesting problem on its own that we plan to investigate in future works. In order to keep the present paper in a reasonable length, we simplify the contract value selection problem by giving some responsibilities to a third party named the grid operator. The role of the grid operator is to constrain the market entry to coalitions able to propose both sufficiently high and sufficiently credible contract values. More formally, let $ \phi \in [0,1] $ be the reliability threshold fixed by the grid operator as a maximum value for the probability of under-producing. The highest contract value that a coalition can propose is thus $ P_{S}^{CRCT \star} $ such that $ Pr[P_{ij} \leq P_{S}^{CRCT \star}] = \phi $. In the Gaussian example, this translates by coalition $ \{i,j\}$ announcing :

@@

\begin{figure}
\includegraphics[scale=.6]{gaussian_P_star}
\caption{$ P_{S}^{CRCT \star} $ depending on reliability parameter $ \phi $ for Gaussian distributions (see equation \ref{eq:P_star}). Blue curve with triangles stands for a coalition S with an expected production of 5 units and a standard deviation of 0.5. Under a grid policy of $ \phi = 0.1 $, it is able to announce a contract value of $ P_{S}^{CRCT \star} = 4.36 $. The same coalition in term of expected production ($\mu = 5 $), but with a higher variance ($ \sigma = 1.5 $, cyan curve with circles) can only afford a smaller contract value of $ P_{S}^{CRCT \star} = 3.07 $. The red curve with diamonds stands for a coalition with a higher expected production $(\mu = 7)$, but with a very high unpredictability ($ \sigma = 5 $). For low values of $ \phi $, this coalition is thus heavily penalized and can only afford a contract of $ 0.59 $ units. Under grid policy $ (\phi = 0.1, P^{MIN} = 2) $, this last coalition is thus not allowed to enter the market (red dot below the horizontal dashed line). }
\label{fig:Gaussian}
\end{figure}

@@

The Gaussian assumption of this small example is convenient as it allows us to write $ P_{S}^{CRCT \star} $ analytically. Nevertheless, such assumption is rather unrealistic in practice. In the following, we keep the same framework but release this Gaussian assumption unless the contrary is specified. This assumption will indeed be convenient for computing some parameter estimates.

@@

In order to get an estimator for $ \alpha $, we solve equation \ref{eq:derivative} in a Gaussian case (as in section \ref{sec:notations}). Furthermore, since considering all the possible interactions between agents is analytically intractable, we use here a mean approximation. Any quantity x that varies over the agent set is thus simplified in its mean value $ \bar{x} $. Solving equation \ref{eq:derivative} for $ \alpha $ in these conditions yields :

@@

Before running the algorithms, we need to calibrate the utility function by choosing the value of the $ \alpha $ parameter. Recall that the purpose of this parameter is to take into account some constraints on the coalition's sizes if needed. In this paper, neither the communication network nor the electrical grid are explicitly considered. Thus, we do not have any technical constraints on coalition sizes even if we designed the utility such that these could be taken into account. We select the desired size as being $ \lfloor N/N_{COAL} \rfloor $ (where $ \lfloor.\rfloor $ means floor). Figure \ref{fig:real_utility2} shows how the mean utility of a coalition evolves with its size when the optimum size is set to 40 agents. Using equation \ref{eq:alpha_star} to estimate $ \alpha $ based on the mean quantities and Gaussian approximations seems to give acceptable results for the utility function behavior.

@@

And we extend this measure to the coalition structures :

@@

We consider that prosumers fail randomly, and we denote by $ \psi \in [0,1] $ the fraction of agents that failed. Figure \ref{fig:resilience} exhibits how the resilience of the coalition structures evolve according to $ \psi $. On the top subplot, $ P^{MIN} $ was voluntarily selected relatively low such that the resiliences of the three structures fit on the same figure. When the $ P^{MIN} $ requirement increases, the differences between the algorithms also increase as visible on the bottom subplot of figure \ref{fig:resilience}. The de-correlated coalitions seem to achieve a more resilient production on the market in the sense that they are able to sustain a higher fraction of node failures.

@@

Because in real situations, agents are prone to failure, resilience is also an important criterion for the quality of the aggregations. We therefore studied how the coalitions are able to remain on the market when their agents fail randomly one by one. We showed that, in this situation, the coalitions resulting from our algorithm better withstand losses of agents.

@@

\begin{itemize}
\item Average wind speed (in $m.s^{-1}$)
\item Nebulosity (integer in $[0,8] $)
\item Temperature (in degree Celsius)
\end{itemize}

@@

Where $ \mathcal{F}_{i}^{heat}(\tau(t), t) $ is the power curve that maps the temperature to a heating consumption, and $ \mathcal{F}_{i}^{elec}(t) $ computes the consumption of agent i (other than heating) at a given hour of the day. In the simulation, all agents have a desired inside temperature $ T_{i} $, supposed to be a constant for simplification. By using thermodynamic laws $ \mathcal{F}_{i}^{heat}(\tau(t), t) $ can be approximated by :
